Log-likelihood for summary data

### 2023-01-20 Working Rmds/MainAnalysis.Rmd
Added option to control whether to do the following: rescale time before fit; fit log-transformed or not. Retained option to fit dose-normalized or not.

```{r}
cvt <- preprocess_data(data.set = cvt_data,
                       names_list =names_list,
                       defaults_list =   defaults_list,
                       ratio_conc_to_dose = 1,
                       calc_loq_factor = 0.45,
                       routes_keep = c("po", "iv"),
                       media_keep = c("blood", "plasma"),
                       impute_loq = TRUE,
                       impute_sd = TRUE,
                       suppress.messages = FALSE)

@@

## Flat model

```{r, run_invivoPKfit_flat}

system.time(
  PK.fit.table.flat <- fit_all(
    data.set = cvt,
    model = "flat",
    modelfun = "analytic",
    preprocess = FALSE,
    optimx_args = list(
      "method" = "bobyqa",
      "itnmax" = 1e6,
      "control" = list("kkt" = FALSE)
    ),
    suppress.messages = TRUE
  )
)

write.csv(PK.fit.table.flat, paste("../inst/ext/PK_fit_table_flat_",
                                    timestamp,
                                    ".csv",
                                    sep=""), row.names = FALSE)

post_flat <- postprocess_data(PK_fit = PK.fit.table.flat,
                               model = "flat")

write.csv(post_flat,
          paste("../inst/ext/PK_fit_postprocess_flat_",
                                    timestamp,
                                    ".csv",
                                    sep=""),
          row.names = FALSE)

rm(PK.fit.table.flat, post_flat)
```

## 1-compartment model

```{r, run_invivoPKfit_1comp}
### do 1-Compartment fit
system.time(
    PK.fit.table.1comp <- fit_all(
      data.set = cvt,
      model = "1compartment",
      modelfun = "analytic",
      preprocess = FALSE,
      optimx_args = list(
                      "method" = "bobyqa",
                      "itnmax" = 1e6,
                      "control" = list("kkt" = FALSE)
                    ),
      suppress.messages = TRUE
  )
)

write.csv(PK.fit.table.1comp,
          paste("../inst/ext/PK_fit_table_1comp_",
                                    timestamp,
                                    ".csv",
                                    sep=""),
          row.names = FALSE)

post_1comp <- postprocess_data(PK_fit = PK.fit.table.1comp,
                               model = "1compartment")

write.csv(post_1comp,
          paste("../inst/ext/PK_fit_postprocess_1comp_",
                                    timestamp,
                                    ".csv",
                                    sep=""),
          row.names = FALSE)

rm(PK.fit.table.1comp, post_1comp)

```

## 2-compartment model

```{r, run_invivoPKfit_2comp}
### do 2-Compartment fit
system.time(
    PK.fit.table.2comp <- fit_all(
      data.set = cvt,
      model = "2compartment",
      modelfun = "analytic",
      preprocess = FALSE,
      optimx_args = list(
                      "method" = "bobyqa",
                      "itnmax" = 1e6,
                      "control" = list("kkt" = FALSE)
                    ),
      suppress.messages = TRUE
    )
)

write.csv(PK.fit.table.2comp, paste("../inst/ext/PK_fit_table_2comp_",
                                    timestamp,
                                    ".csv",
                                    sep=""), row.names = FALSE)

post_2comp <- postprocess_data(PK_fit = PK.fit.table.2comp,
                               model = "2compartment")

write.csv(post_2comp,
          paste("../inst/ext/PK_fit_postprocess_2comp_",
                                    timestamp,
                                    ".csv",
                                    sep=""),
          row.names = FALSE)

rm(PK.fit.table.2comp, post_2comp)
```

# Test case

DTXSID1022265 (Alachlor) is an example of one chemical where there is only one Reference ID, but the PO and IV data are on totally different scales -- IV concentration/dose ranges from 1e-3 to 1e-1, but PO concentration/dose ranges from 1e-5 to 1e-4. Now that we are fitting on natural scale, this means that with reference-specific sigma, the curve really only fits the IV data, because sigma is so large compared to the PO data scale. This results in a too-small Fgutabs estimate that does not fit the peak very well. But with "study" specific sigma -- chemical + species + reference + route + media -- the PO sigma can be on a different order than the IV sigma, and is.

Another possible thing to try is to go back to fitting on the log scale. I had switched to the natural scale because I wasn't sure how to derive the "summary" parameterization of the residual distribution for the log scale. But I could try again.

>>>>

```{r}
cvt <- preprocess_data(data.set = cvt_data,
                       names_list =names_list,
                       defaults_list =   defaults_list,
                       ratio_conc_to_dose = 1,
                       calc_loq_factor = 0.45,
                       routes_keep = c("po", "iv"),
                       media_keep = c("blood", "plasma"),
                       impute_loq = TRUE,
                       impute_sd = TRUE,
                       study_def = c("DTXSID", "Species", "Reference", "Route", "Media"),
                       suppress.messages = TRUE)

@@

```{r}
for(this_rescale_time in c(FALSE, TRUE)){
  for(this_fit_conc_dose in c(FALSE, TRUE)){
    for(this_fit_log_conc in c(FALSE, TRUE)){
      if(!(this_rescale_time %in% FALSE &
           this_fit_conc_dose %in% FALSE &
           this_fit_log_conc %in% FALSE)){ #already did the all FALSE case
      for (this_model in c("flat",
                           "1compartment",
                           "2compartment")){
        print(paste0("model = ", this_model, "\n",
                     "fit_log_conc = ", this_fit_log_conc, "\n",
                     "fit_conc_dose = ", this_fit_conc_dose, "\n",
                     "rescale_time = ", this_rescale_time))
        system.time(
          PK.fit.table <- fit_all(
            data.set = cvt,
            model = this_model,
            modelfun = "analytic",
            preprocess = FALSE,
            fit_conc_dose = this_fit_conc_dose,
            fit_log_conc = this_fit_log_conc,
            rescale_time = this_rescale_time,
            optimx_args = list(
              "method" = "bobyqa",
              "itnmax" = 1e6,
              "control" = list("kkt" = FALSE)
            ),
            suppress.messages = FALSE
          )
        )

        write.csv(PK.fit.table,
                  paste("../inst/ext/PK_fit_table_",
                        this_model,
                        "_fit_log_conc_",
                        this_fit_log_conc, "_",
                        "_fit_conc_dose_",
                        this_fit_conc_dose, "_",
                        "_rescale_time_",
                        this_rescale_time, "_",
                        timestamp,
                        ".csv",
                        sep=""), row.names = FALSE)

        post_table <- postprocess_data(PK_fit = PK.fit.table,
                                       model= this_model)

        rm(PK,fit.table)

        write.csv(post_table,
                  paste("../inst/ext/PK_fit_postprocess_",
                        this_model,
                        "_fit_log_conc_",
                        this_fit_log_conc, "_",
                        "_fit_conc_dose_",
                        this_fit_conc_dose, "_",
                        "_rescale_time_",
                        this_rescale_time, "_",
                        timestamp,
                        ".csv",
                        sep=""), row.names = FALSE)

        rm(post_table)
      } #end loop over models
      } #end if not all FALSE
    } #end loop over this_fit_log_conc
  } #end loop over this_fit_conc_dose
} #end loop over this_rescale_time
```

### 2023-01-20 Working Rmds/summary_loglike_lognormal.Rmd
Added option to control whether to do the following: rescale time before fit; fit log-transformed or not. Retained option to fit dose-normalized or not.

$$ L_{ij} = \frac{1}{\sigma \sqrt{2 \pi}} \exp \left[ \frac{-1}{2} \left( \frac{y_{ij} - \mu_i}{\sigma} \right) ^2 \right] $$

@@

$$LL_{ij} = \log \left( \frac{1}{\sigma \sqrt{2 \pi}} \right) + \frac{-1}{2} \left( \frac{y_{ij} - \mu_i}{\sigma} \right) ^2 $$
## Total log-likelihood of group i

@@

$$ LL_i = \sum_{j=1}^{M_i} LL_{ij} =  \sum_{j=1}^{M_i} \left[ \log \left( \frac{1}{\sigma \sqrt{2 \pi}} \right) + \frac{-1}{2} \left( \frac{y_{ij} - \mu_i}{\sigma} \right) ^2 \right]$$
Breaking apart the two terms inside the summation (square brackets), and expanding the square in the second term, this becomes

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \frac{-1}{2 \sigma^2} \left[\sum_{j=1}^{M_i} \left( y_{ij}^2 - 2 \mu_i y_{ij} + \mu_i^2  \right) \right]  $$
Further splitting apart the terms inside the summation (square brackets), this becomes

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \sum_{j=1}^{M_i} y_{ij}^2 + \\
\frac{2 \mu_i}{2 \sigma^2} \sum_{j=1}^{M_i} y_{ij} + \\
\frac{-\mu_i^2 M_i}{2 \sigma^2} $$

Substitute in the definition of the sample mean:

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \sum_{j=1}^{M_i} y_{ij}^2 + \\
\frac{2 \mu_i}{2 \sigma^2} M_i \bar{y}_i + \\
\frac{-\mu_i^2 M_i}{2 \sigma^2} $$

Using the definitions of the sample variance and sample mean, it can be shown (see the Appendix) that

$$ \sum_{j=1}^{M_i}  y_{ij}^2 = \left(M_i - 1 \right)  s_i^2 + M_i \bar{y}_i^2  $$
Substitute this back into the log-likelihood:

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \left[ \left(M_i - 1 \right)  s_i^2 + M_i \bar{y}_i^2 \right] + \\
\frac{2 \mu_i}{2 \sigma^2} M_i \bar{y}_i + \\
\frac{-\mu_i^2 M_i}{2 \sigma^2} $$

This simplifies to

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \left( \\
\left(M_i - 1 \right)  s_i^2 + M_i \bar{y}_i^2  + \\
-2 \mu_i M_i \bar{y}_i + \\
\mu_i^2 M_i \\
\right) $$

which further simplifies to

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \left( \\
\left(M_i - 1 \right)  s_i^2 + \\
M_i \left( \bar{y}_i - \mu_i \right)^2 \\
\right) $$

>>>>

$$ L_{ij} = \frac{1}{\sigma \sqrt{2 \pi}} \exp \left[ \frac{-1}{2} \left( \frac{\log y_{ij} -  \mu_i}{\sigma} \right) ^2 \right] $$

@@

$$LL_{ij} = \log \left( \frac{1}{\sigma \sqrt{2 \pi}} \right) + \frac{-1}{2} \left( \frac{\log y_{ij} -  \mu_i}{\sigma} \right) ^2 $$
## Total log-likelihood of group i

@@

$$ LL_i = \sum_{j=1}^{M_i} LL_{ij} =  \sum_{j=1}^{M_i} \left[ \log \left( \frac{1}{\sigma \sqrt{2 \pi}} \right) + \frac{-1}{2} \left( \frac{\log y_{ij} - \mu_i}{\sigma} \right) ^2 \right]$$
Breaking apart the two terms inside the summation (square brackets), and expanding the square in the second term, this becomes

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \frac{-1}{2 \sigma^2} \left[\sum_{j=1}^{M_i} \left( \left( \log y_{ij} \right) ^2 - 2  \mu_i \log y_{ij} + \mu_i^2  \right) \right] $$
Further splitting apart the terms inside the summation (square brackets), this becomes

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \sum_{j=1}^{M_i} \left( \log y_{ij} \right)^2 + \\
\frac{2 \mu_i}{2 \sigma^2} \sum_{j=1}^{M_i} \log y_{ij} + \\
\frac{-\mu_i^2 M_i}{2 \sigma^2} $$

Here is where things get complicated with a log-normal distribution. With a normal distribution, we could substitute in the definition of the sample mean for $\sum y_{ij}$. With a log-normal distribution, we instead have $\sum \log y_{ij}$, which does not correspond to the definition of the sample mean. At best we could do the log of the sample mean -- but that doesn't give us what we are looking for.

$$\log \bar{y}_i = \log \sum_{j=1}^{M_i} y_{ij} \neq \sum_{j=1}^{M_i} \log y_{ij}$$
If log-scale population mean is $m_i$ and log-scale standard deviation is $s_i$, and sample mean is $\bar{y}_i$ and sample standard deviation is $S_i$, then

$$ m_i = \log \frac{\bar{y}_i^2}{\sqrt{S_i^2 + \bar{y}_i^2}} = \frac{1}{M_{i}} \sum_{j=1}^{M_i} \log y_{ij}$$
and

$$ s_i = \sqrt{ \log \left( 1 + \frac{S_i^2}{\bar{y}_i^2} \right)} = \sqrt{ \frac{1}{M_i - 1} \sum_{i=1}^{M_i} (\log y_{ij} - m_i)^2}$$

Substitute in the definition of the log-scale mean for $\sum_{j=1}^{M_i} \log y_{ij}$

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \sum_{j=1}^{M_i} \log y_{ij}^2 + \\
\frac{2 \mu_i}{2 \sigma^2} M_i \log \frac{\bar{y}_i^2}{\sqrt{S_i^2 + \bar{y}_i^2}} + \\
\frac{-\mu_i^2 M_i}{2 \sigma^2} $$

Using the definitions of the log-scale sample variance and log-scale sample mean, it can be shown that

$$ \sum_{j=1}^{M_i}  \left( \log y_{ij} \right)^2 = \left(M_i - 1 \right)  s_i^2 + M_i m_i^2 = (M_i -1) \log \left( 1 + \frac{S_i^2}{\bar{y}_i^2} \right) + M_i  \left(\log \frac{\bar{y}_i^2}{\sqrt{S_i^2 + \bar{y}_i^2}}\right)^2 $$
Substitute this back into the log-likelihood:

$$ LL_i = M_i \log \left(\frac{1}{\sigma \sqrt{2 \pi}} \right) + \\
\frac{-1}{2 \sigma^2} \left( (M_i -1) \log \left( 1 + \frac{S_i^2}{\bar{y}_i^2} \right) + M_i  \left(\log \frac{\bar{y}_i^2}{\sqrt{S_i^2 + \bar{y}_i^2}}\right)^2 \right) + \\
\frac{2  \mu_i}{2 \sigma^2} M_i \log \frac{\bar{y}_i^2}{\sqrt{S_i^2 + \bar{y}_i^2}} + \\
\frac{-\mu_i^2 M_i}{2 \sigma^2} $$

This looks horrible, but I think it is correct.
